---
title: "Additional technical details"
author: "sstools Team"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Additional technical details}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 6,
  fig.height = 6
)
```

## Small sample bias

The ssdtools package uses the method of Maximum Likelihood (ML) to estimate parameters for each distribution
that is fit to the data. Statistical theory says that maximum likelihood estimators are asymptotically
unbiased, but does not guarantee performance in small samples. A detailed account of the issue of small sample bias in estimates can be found in the following[pdf](https://github.com/bcgov/ssdtools/blob/master/vignettes/small-sample-bias.pdf).

## Numerical instability of the Burr III

## Numerical properties of the inverse Pareto distribution

## Parameter estimates using L-moments (Burr III)

## Closed form estimation (Burr III)

## The ssd_fit_burrlioz function in ssdtools




```{r, results = "asis", echo = FALSE}
cat(ssdtools::licensing_md())
```
